---
title: "Class 26: Assignment"
author: "Hughes/Fisher"
date: "Apr 29 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
```


### Multiple Logistic Regression Example: Contraceptive Use

The dataset `contraceptiveuse.csv` contains information from $n$ = 1607 currently married and fecund women interviewed in the Fiji Fertility Survey, according to age, education, desire for more children and current use of contraception.

```{r}
c.use <- read.csv("contraceptiveuse.csv")
kable(c.use)
```

We want to build a good logistic regression model for predicting $p = Pr(\textrm{using contraceptives})$ as a function of age, education and desire for more children.  Because we have three predictor variables to consider, this is a **multiple logistic regression** model.  Since all three predictors are categorical variables here, and coded as such, they are treated automatically as factors by R.   

-------------------------------------

**QUESTION 1.** Fit the full main effects multiple logistic regression model using `age`, `education`, and `wantsMore` as the predictor variables. Run a full-model $\chi^2$ likelihood ratio test to see if this model is useful for predicting $p = Pr(\textrm{using contraceptives})$.

```{r}
c.use.fit <- glm(cbind(using, notUsing) ~ wantsMore + education + age, data=c.use, family=binomial(link=logit))
drop1(c.use.fit, test="Chisq")
```

*ANSWER HERE: We can see that this model is useful for predicting as all the predictors are significant. `wantsMore` is a significant predictor of contraceptive use ($\chi^2$ = 50.501, $df$ = 1, $p$-value = 1.191 x 10^-12^). `education is a significant predictor of contraceptive use ($\chi^2$ = 6.971, $df$ = 1, $p$-value = 0.008286). Finally, `age` is a significant predictor of contraceptive use ($\chi^2$ = 43.948, $df$ = 3, $p$-value = 1.548 x 10^-9^)*


-------------------------------------

**QUESTION 2.** Run the `summary` function on your main effects model from part 1. Provide an interpretation, in context, of the model's intercept estimate but **expressed in odds terms**; i.e. calculate $exp(\hat{\beta_0})$ and interpret the value.

```{r}
summary(c.use.fit)
exp(coef(c.use.fit)[1])
```

*ANSWER HERE: The odds of using a contraceptive are 0.446.* 


-------------------------------------

**QUESTION 3.** Look at the `summary` output from part 2.  Provide an interpretation for each of the five estimated $\beta$-coefficients for the predictors under study.  (Remember, everything is in terms of log odds at present, so you need to exponentiate the estimates to aid interpretation of the size of the effects in terms of odds.)

```{r}
exp(coef(c.use.fit)[2])
exp(coef(c.use.fit)[3])
exp(coef(c.use.fit)[4])
exp(coef(c.use.fit)[5])
exp(coef(c.use.fit)[6])
```

*ANSWER HERE: The odds of using a contraceptive and wanting more children is 0.434 times that of not using a contraceptive and wanting more children. The odds of using a contraceptive and having low education is 0.723 times that of not using a contraceptive and having high education. The odds of using a contraceptive and being aged 25-29 is 0.1.476 times that of not using a contraceptive and not being aged 25-29. The odds of using a contraceptive and being aged 30-39 is 2.481 times that of not using a contraceptive and not being aged 30-39. The odds of using a contraceptive and being aged 40-49 is 3.285 times that of not using a contraceptive and not being aged 40-49.*


-------------------------------------

**QUESTION 4.** Which of the three predictors are significant predictors of contraceptive use?

*ANSWER HERE: All three predictors are significant predictors of contraceptive use (education, age, and wanting more children).* 


-------------------------------------

**QUESTION 5.** Find a 95% confidence interval for the odds of using contraceptives between women who want more children vs those who do not want more children (regardless of age or education level).

```{r}
c.use.fit.children <- glm(cbind(using, notUsing) ~ wantsMore, data=c.use, family=binomial(link=logit))
exp(confint(c.use.fit.children))
```

*ANSWER HERE: We can be 95% confident that women who want more children are 0.282 to 0.435 times as likely to use contraceptives than women who don't want more children.*  


-------------------------------------

**QUESTION 6.** Find the predicted probabilities of using contraceptives for each of the 16 combinations of `age`, `education`, and `wantsMore`.  Use `cbind` to combine these predicted probabilities back with the original data set.  According to your model, which type(s) of women are generally most likely to use contraceptives?

```{r}
pred <- predict(c.use.fit, type="response")
together <- cbind(c.use, pred)
kable(together)
```

*ANSWER HERE: Women in their 40s who don't want more children.* 